---
title: "Setting up docker on server"
output:
  html_document: default
  pdf_document: default
---

## Stopping and starting all running containers

```{bash,eval=FALSE}
docker stop $(docker ps -a -q)
docker rm $(docker ps -a -q)
```

## Pulling and running my pre-built images


The RStudio container has been altered to make rstudio a sudo user. Postgis has been installed in order to get shp2psql and raster2psql. Libgdal-dev has been added along with unixodbc. PLR has been added to the postgis image.

The flags set the data valumes to folders in the home directory.

```{bash,eval=FALSE}
## Note the data directories are given completely free permsions. No security here! However none is really needed and this prevents problems with users finding they are denied permision to do things. 

mkdir -p /home/duncan/postgres_data
chmod 777 -R /home/duncan/postgres_data
mkdir -p /home/duncan/rstudio_server
chmod 777 -R /home/duncan/rstudio_server

docker run --name "postgis-plr" -p 25432:5432 -d -v /home/duncan/postgres_data:/var/lib/postgresql dgolicher/postgis-plr

docker run --name "rstudio" --link postgis-plr:postgis -d -p 8788:8787 -v /home/duncan/rstudio_server:/home/rstudio dgolicher/rstudio
```

R runs from http://172.16.49.31:8788

The default postgis user is docker with pwd docker

## Using .pgpass

In order to run commands within R studio without being asked for a password for the default user (docker) with password docker needs a simple .pgpass file to be placed in the home directory of the user making the calls.

```{bash,eval=FALSE}
hostname:port:database:username:password

i.e

postgis:*:*:docker:docker
```

Save this as simply .pgpass
The permissions need to be set to u=rw (0600) or less so go into the shell and change this with sudo chmod 0600 .pgpass

## Creating data bases from R

```{r}
nm<-"brant"
com<-paste("dropdb -h postgis -U docker ",nm,sep="")
com
system(com)
com<-paste("createdb -h postgis -U docker ",nm,sep="")
com
system(com)
```


```{r}

library(RODBC)
con<-odbcConnect("brant")
str(con)
odbcQuery(con,"create extension postgis")
odbcQuery(con,"create extension plr")
```


```{r}
library(raster)
library(rgdal)
r<-raster("shiny_morph/cold_bay_3857_clip.tiff")
plot(r)
(r@extent@xmax-r@extent@xmin)/r@ncols
(r@extent@ymax-r@extent@ymin)/r@nrows

```


```{r}

r_nm<-r@file@name
r_nm
grid_side=10
command <- paste("raster2pgsql -d  -M  ",r_nm, " -F -t ",grid_side,"x",grid_side," tmp|psql -h postgis -U docker -d brant",sep="")
system(command)

query<-"CREATE OR REPLACE FUNCTION median (float[]) RETURNS float AS '
x<-arg1
x<-as.numeric(as.character(x))
x<-na.omit(x)
median(x)'
LANGUAGE 'plr' STRICT;
CREATE OR REPLACE FUNCTION q10 (float[]) RETURNS float AS '
x<-arg1
x<-as.numeric(as.character(x))
x<-na.omit(x)
quantile(x,0.1,na.rm=TRUE)'
LANGUAGE 'plr' STRICT;
CREATE OR REPLACE FUNCTION q90 (float[]) RETURNS float AS '
x<-arg1
x<-as.numeric(as.character(x))
x<-na.omit(x)
quantile(x,0.9,na.rm=TRUE)'
LANGUAGE 'plr' STRICT;"

odbcQuery(con,query)

query<-"drop table if exists tmp2;
create table tmp2 as
select rid, st_envelope(rast) geom,
q10((st_dumpvalues(rast)).valarray) q10,
median((st_dumpvalues(rast)).valarray) median,
q90((st_dumpvalues(rast)).valarray) q90
from tmp"

odbcQuery(con,query)


```


```{r}

getquery <- function(query) {
    query <- paste("create view temp_view as ", query, sep = "")
    odbcQuery(con, query)
    dsn<-"PG:dbname='brant' host='postgis' port=5432 user= 'docker'"
    result <- readOGR(dsn, "temp_view")
    odbcQuery(con, "drop view temp_view")
    return(result)
}
grd<-getquery("select * from tmp2 where q10<0")
plot(grd)
```


## Example query that might be adapted later to extract resources onto grid

```{sql eval=FALSE}
select max(t.rid),min(t.rid),t2.rid,min(t2.median),median((st_dumpvalues(st_union(st_clip(rast,geom)))).valarray) med
from tmp t,
(select * from tmp2) t2
where st_intersects(rast,geom)
group by t2.rid

```


